D3064R0: OOTA Execution on Real Implementations is Provably Vacuous

### 2023-12-15 ootavacuous.tex
Add appendix on coe, fre, and rfe IPC modes
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\begin{document}
\title{D3064R0: OOTA Execution is Provably Vacuous}

@@

Third, one could imagine implementations whose analysis is restricted to
a thread.
In this case, the analysis stops at loads to objects that were last
stored to by other threads (``rfe''), stores to objects where the next
and/or previous stores in that object's modification order were executed
by other threads (``coe'')), and loads from objects for which the value
returned by a given load was overwritten by some other thread (``fre'').

@@

Carrying out this type of optimization requires that the implementation
classify dependencies as semantic or not.
Given the limited resources available, almost all implementations will
choose to approximate sdep, which entails some misclassifications, with
the consequences summarized in
Table~\ref{tab:Semantic-Dependency Classification}.
The first row of the table depicts the case where the implementation
correctly identifies a semantic dependency, and thus exhibits temporal
behavior for the execution at hand of the corresponding code.
The second row depicts the case where the implementation misclassifies
the dependency as non-semantic, which will result in an incorrect
execution.
In this case, the proper course of action is to fix that bug.
The third row depicts the case where the implementation misclassifies
the dependency as semantic, thus missing a chance to optimize.
The resulting execution is nevertheless correct, and is also temporal.
The fourth and final row depicts the case where the implementation
correctly classifies the dependency as non-semantic, in which case
the execution is both correct and atemporal.

@@

@@

Nonetheless, all errors and omissions in this paper are the sole property
of the authors, and the appearance of a name in this paragraph does
not in any way constitute agreement with anything in this paper.

>>>>

\begin{document}
\title{D3064R0: OOTA Execution on Real Implementations is Provably Vacuous}

@@

Third, one could imagine implementations whose analysis is restricted to
a thread.
In this case, the analysis stops at loads to objects that were last
stored to by other threads (``rfe''), stores to objects where the next
and/or previous stores in that object's modification order were executed
by other threads (``coe'')), and loads from objects for which the value
returned by a given load was overwritten by some other thread
(``fre'').\footnote{
	See Appendix~\ref{app:Inter-Process Communications} for more
	information on these memory-reference interprocess communications
	modes.}

@@

Carrying out this type of optimization requires that the implementation
classify dependencies as maybe semantic or definitely not.
Given the limited resources available, almost all implementations will
choose to approximate sdep, which entails some misclassifications, with
the consequences summarized in
Table~\ref{tab:Semantic-Dependency Classification}.
The first row of the table depicts the case where the implementation
correctly identifies a semantic dependency, and thus exhibits temporal
behavior for the execution at hand of the corresponding code.
The second row depicts the case where the implementation misclassifies
the dependency as non-semantic, which will result in an incorrect
execution.
In this case, the proper course of action is to fix that bug.
The third row depicts the case where the implementation misclassifies
the dependency as semantic, thus missing a chance to optimize.
The resulting execution is nevertheless correct, and is also temporal.
The fourth and final row depicts the case where the implementation
correctly classifies the dependency as non-semantic, in which case
the execution is both correct and atemporal.

@@

\section{Inter-Process Communications}
\label{app:Inter-Process Communications}

This section more clearly defines the ``coe'', ``fre'', and ``rfe''
types of interprocess communication (also collectively called ``links'')
and gives evidence on the temporal qualities of these different types.

First, context.
The final ``e'' in all three acronyms stands for ``external'' as opposed
to within-thread communication (``coi'', ``fri'', and ``rfi'') and as
also opposed to all communication, whether inter-thread or intra-thread
(``co'', ``fr'', and ``rf'').

Next, definitions:

\begin{description}
\item[coe]
	Coherency (external), which connects a store to any other store
	to that same object that overwrites either the first store's
	value or that overwrites some later value.
\item[fre]
	From-read (external), which connects a load to any other store
	object that overwrites either the value that the load returned
	or that overwrites some later value.
\item[rfe]
	Reads-from (external), which connects a store to any load from
	that same object that returns the value stored.
\end{description}

\begin{figure}[tb]
\begin{center}
\resizebox{4in}{!}{\includegraphics{ipc}}
\caption{IPC Diagram for coe, fre, and rfe}
\label{fig:IPC Diagram for coe, fre, and rfe}
\end{center}
\end{figure}

Thirdly, Figure~\ref{fig:IPC Diagram for coe, fre, and rfe} illustrates
the coe, fre, fri, and rfe links for four threads with time advancing
from the top to the bottom of the figure.  Starting with coe, even though
Thread~1's store was later in global time than that of Thread~0, that of
Thread~1 came first in coe order, followed by those of Threads~2 and ~3.
There is thus a coe link from Thread~1's store to every other store,
from Thread~0's store to those of Threads~2 and~3, and from Thread~2's
store to that of Thread~3.

Thread~3's load reads the value stored by Thread~0's store, so there is
an rfe link from that store to that load.  There is also an fre link from
that load to Thread~2's store and an fri link to Thread~3's later store.
Note that this fre link goes backwards in time.

Finally, there might be a desire for hard evidence that coe and fre
links really can go backwards in time.
We provide this evidence on x86 to demonstrate that these effects
are not confined to weakly ordered systems.
This system is a laptop with Intel(R) Core(TM) i9-10885H CPU @ 2.40GHz,
having a single socket with 8~cores, each having a pair of hardware
threads.
The code generating this data may be found in the \path{CodeSamples/cpu}
directory of the
perfbook~\cite{McKenney2018ParallelProgramming-2018-12-08a}.\footnote{
	\co{git clone git://git.kernel.org/pub/scm/linux/kernel/git/paulmck/perfbook.git}}

\begin{figure}[tb]
\begin{center}
\resizebox{3in}{!}{\includegraphics{coe-sh-out}}
\caption{On x86, coe Links Are Atemporal}
\label{fig:On x86, coe Links Are Atemporal}
\end{center}
\end{figure}

Figure~\ref{fig:On x86, coe Links Are Atemporal}
shows the results of 15 of the laptop's hardware threads simultaneously
storing their identifying integer (ranging from 0 to 14) to a shared
variable, then repeatedly doing timestamped reads from that variable.
The x-axis displays time in TSC periods, each of which is about 0.417~ns
in duration.
The y-axis shows the number of distinct opinions that the hardware
threads have as a function of time, a number that is frequently rather
larger than one.
The backwards-in-time coe link from Thread~1 to Thread~0 in
Figure~\ref{fig:IPC Diagram for coe, fre, and rfe}
is therefore entirely plausible.

\begin{figure}[tb]
\begin{center}
\resizebox{5in}{!}{\includegraphics{coe-out-4}}
\caption{On x86, coe Links Are Partially Ordered}
\label{fig:On x86, coe Links Are Partially Ordered}
\end{center}
\end{figure}

\begin{figure}[tb]
\begin{center}
\resizebox{3in}{!}{\rotatebox{90}{\includegraphics{store15tred}}}
\caption{On Power5, coe Links Are Also Partially Ordered}
\label{fig:On Power5, coe Links Are Also Partially Ordered}
\end{center}
\end{figure}

However, Figure~\ref{fig:On x86, coe Links Are Partially Ordered}
shows that the order of reads of each value at each thread are
consistent with a number of global orders, as required.
Figure~\ref{fig:On Power5, coe Links Are Also Partially Ordered}
shows a rather more elaborate partial order from a Power5
system~\cite{McKenney2018ParallelProgramming-2018-12-08a}.

\begin{figure}[tb]
\begin{center}
\resizebox{3in}{!}{\includegraphics{fre-sh-out}}
\caption{On x86, fre Links Are Atemporal}
\label{fig:On x86, fre Links Are Atemporal}
\end{center}
\end{figure}

Figure~\ref{fig:On x86, fre Links Are Atemporal}
plots a histogram of the elapsed time from the last load to return the
old value to the store that provided the new value.
Note well that most of the data falls into negative time, that is to say,
the last load of the old value executes about 50 timestamp periods (about
20~nanoseconds) \emph{after} the store that overwrote that same old value.
That is, fre links can and do go backwards in time.
This corresponds to the fre link between Thread~3 an Thread~2 depicted in
Figure~\ref{fig:IPC Diagram for coe, fre, and rfe}.

\begin{figure}[tb]
\begin{center}
\resizebox{3in}{!}{\includegraphics{rfe-sh-out}}
\caption{On x86, rfe Links Are Temporal}
\label{fig:On x86, rfe Links Are Temporal}
\end{center}
\end{figure}

Figure~\ref{fig:On x86, rfe Links Are Temporal}
plots a histogram of the elapsed time from a store of a new value
to the first load of that new value.
Note well that all of the data falls into positive time, indicating
that rfe links always go forward in time, as those familiar with
computer hardware and the laws of physics would expect.

And this is exactly why the C++ memory model guarantees ordering from
rfe links but not the coe and fre links for relaxed accesses in the
absence of other ordering from stronger atomic memory accesses or
\co{atomic_thread_fence()}.

@@

Nonetheless, all errors and omissions in this paper are the sole property
of the authors, and the appearance of a name in this appendix does
not in any way constitute agreement with anything in this paper.
